---
title: "Datanalysis for the Trachypithecus Data"
author: Detlef Groth and Truong Van Nguyen
date: 2026-05-29
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Datanalysis for the Trachypithecus Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
include-before: |
    <style> 
    body { max-width: 1000px; font-family: Candara, sans-serif;} 
    table { min-width: 400px; border-spacing: 5px;  border-collapse: collapse; }
    </style>
---

```{r echo=FALSE,results="hide"}
library(knitr)
```

## Introduction

This is the vignette for the package [trachy](https://github.com/mittelmark/trachy) which contains the data and the code
for the paper of Truong et. al. (2025). DNA metabarcoding reveals dietary adaptations of Trachypithecus langurs in limestone and rainforest habitats. (in preparation)

The code below was used to create the figures for the statistical analysis which were submitted with the written text.

## Data preparation

Let's first load and show the data which are part of the  package.  The can be
loaded using the `trachy$tdata` method. We start with the meta data.

```{r}
library(trachy)
if (!require("sbi")) {
    install.packages("https://github.com/mittelmark/sbi/releases/download/v0.1.0/sbi_0.1.0.tar.gz",repos=NULL)
}
library(sbi)
meta=trachy$tdata('meta')
head(meta)
```

Below a summary of the samples:

```{r}
with(meta,ftable(species,habitat,seasonal))
```

## Family Data Analyis

Let's now load the raw family data:

```{r}
options(width=120)
fmly=trachy$tdata('family')
dim(fmly)
fmly[1:4,1:10]
```

Let's merge both tables now into one to attach season and habitat  information
to the data set:


```{r}
mdata=merge(fmly,unique(meta[,c('habitat','code','seasonal')]))
dim(mdata)
mdata[1:8,1:8]
```

Let's reorder the columns so that we have meta information first and the NGS data.


### Season Medians

Let's calculate the season means. For this we first merge the data:

```{r}
agg=with(mdata,aggregate(mdata[,6:(ncol(mdata))],
    by=list(seasonal,species),median))
colnames(agg)[1]="Season"
colnames(agg)[2]="Species"
agg[,1:10]
```

Let's tabulate this:

```{r results="asis"}
library(knitr)
aggr=agg[,1:12]
aggr[,3:12]=round(aggr[,3:12],1)
aggr[,2]=gsub(" ","",aggr[,2])
colnames(aggr)=substr(colnames(aggr),1,6)
colnames(aggr)[1]="Season"
colnames(aggr)[2]="Species"
knitr::kable(aggr)
```


### Visualization of Plant family similarities using MDS


```{r fig=TRUE,fig.width=8,fig.height=8,out.width=800}
par(mai=c(0.8,0.8,0.5,0.2))
M=t(as.matrix(mdata[,6:ncol(mdata)]))
dim(M)
M[1:6,1:5]
col=rep("black",nrow(M))
col[1:5]="red"
rownames(M)=substr(rownames(M),1,4)
trachy$mds_plot(M,method="spearman",col.labels=col,grid=FALSE)
abbrev.df=data.frame(abbrev=substr(colnames(mdata)[6:(ncol(mdata))],1,4),
    family=colnames(mdata)[6:(ncol(mdata))])
```

**Figure  MDS of plant  families:**  Multidimensional  Scaling of the Spearman
correlation  based  distances ( D=(1-rho)/2 ) for the plant families. The five
most abundant  species,  Moraceae  (Mor),  Fagaceae (Fag),  Sapotaceae  (Sap),
Fabaceae   (Fab)  and  Rubiaceae   (Rub)  are  shown  in  red.  Full  list  of
abbreviations is given in the supplementary materials.

### MDS of samples
        

We will now use the aggregated data.

```{r fig.cap="MDS with Spearman correlation as distance measure",fig.width=8,fig.height=8,out.width=800}
species=list(vmt="Tdel",vht="That",vb="Tger",vx="Tcre")
colors=list(Tger="#dd7733",Tdel=6,Tcre=3,That=4)
spec=gsub("_.+","",mdata$code)
head(spec)
spec = unlist(species[spec])
cols=unlist(colors[spec])
head(cols)
pchs=c(18,20)[as.numeric(as.factor(mdata$seasonal))]
S=as.matrix(mdata[,6:ncol(mdata)])
dim(S)
trachy$mds_plot(S,
    method="spearman",points=TRUE,pch=pchs,cex=3,col=cols,row.labels=FALSE,grid=FALSE)
legend('bottomleft',legend=names(colors)[c(2,4,1,3)],fill=unlist(colors)[c(2,4,1,3)],cex=1.2)
legend('bottomright',legend=c("Dry","Wet"),pch=c(18,20),cex=1.2,pt.cex=2)
```

**Figure  MDS of Trachypithecus samples:**  Multidimensional  Scaling of the Spearman
correlation  based  distances ( D=(1-rho)/2 ) for the average values for each month based of the plant family data.  Every species has up to 12 datapoints, 

### Averages of the MDS

```{r fig=TRUE,fig.width=8,fig.height=8,out.width=800}
trachy$mds_plot(S,
    method="spearman",points=TRUE,
    pch=15,cex=3.2,col=cols,row.labels=FALSE,grid=FALSE,group.mean=TRUE)
legend('bottomleft',legend=names(colors)[c(2,4,1,3)],fill=unlist(colors)[c(2,4,1,3)],cex=1.2)
```


### MDS plot for every species

```{r fig=TRUE,fig.width=10,fig.height=10,out.width=1000}
par(mfrow=c(2,2),mai=c(0.7,0.7,0.6,0.2),omi=c(0.5,0.2,0.1,0.1))
options(warning=-1)
sp = unique(agg$Species)
sp = sp[c(2,4,3,1)]
x = 0
for (species in sp) {
    x=x+1
    mdat=mdata[mdata$species==species,6:ncol(mdata)]
    idx=which(apply(mdat,2,sum)>5)
    mdat=mdat[,idx]
    a=agg[agg$Species==species,colnames(mdat)]
    colnames(mdat)=substr(colnames(mdat),1,4)
    lf=log2(a[1,]/a[2,])
    col=rep("black",length(lf))
    col[lf < -1] = "blue"
    col[lf >  1] = "orange"
    xlab="Dim 1"
    ylab="Dim 2"
    if (x %in% c(1,2)) {
        xlab=""
    }
    if (x %in% c(2,4)) {
        ylab=""
    }
    if (x == 1) {
        par(mai=c(0.5,0.7,0.6,0.1))
    }
    if (x == 2) {
        par(mai=c(0.5,0.6,0.6,0.2))
    }
    if (x == 3) {
        par(mai=c(0.7,0.7,0.4,0.1))
    }
    if (x == 4) {
        par(mai=c(0.7,0.6,0.4,0.2))
    }
    trachy$mds_plot(t(mdat),method="spearman",grid=FALSE,main=species,col.labels=col,
        xlab=xlab,ylab=ylab)
}
trachy$bottom_legend(
    labels=c("dry","same","wet"),col=c("orange","black","blue"),cex=1.5)
```
**Figure  MDS of plant  families foreach Trachypithecus species:**  Multidimensional  Scaling of the Spearman
correlation  based  distances ( D=(1-rho)/2 ) forthe plant families.  Families
where the  log2-changes  between  wet and dry season are lower than -1 (plants
found more often in wet  season)  are shown in blue color, and plant  families
found  more  often in dry  season  (log2-foldchange  > 1) are shown in orange.
Families not having a log2-foldchange between -1 and 1 are shown in black.

### PCA

```{r fig.width=10,fig.height=10,fig.cap="PCA on plant family abundance data"}
pca.plot = function (pca,...) {
    pcasum=summary(pca)$importance[2,] * 100
    plot(pca$x,...,type="n",xlab=sprintf("PC1 (%.1f%%)",pcasum[1]),
        ylab=sprintf("PC2 (%.1f%%)",pcasum[2]))
    #grid()
    points(pca$x,...)
}
par(mfrow=c(2,2))
pca=prcomp(mdata[,6:(ncol(mdata))])
pca.plot(pca,col=cols,pch=pchs,cex=3,main="raw data")
pca=prcomp(scale(mdata[,6:(ncol(mdata))]))
pca.plot(pca,col=cols,pch=pchs,cex=3,main="scaled data")
pca=prcomp(scale(log2(mdata[,6:(ncol(mdata))]+1)))
pca.plot(pca,col=cols,pch=pchs,cex=3,main="scaled log-data")
pca=prcomp(log2(mdata[,6:(ncol(mdata))]+1))
pca.plot(pca,col=cols[spec],pch=pchs,cex=3,main="log-data")
```

**Figure: PCA of plant family abundance  data.** Shown are the results for the
raw, data, z-score scaled data, log2+1 transformed z-score data and log2-data.
Regardless of the data normalization all plots show a clear species separation
and a less  obvious  separation  of the  seasons  for each  species.  The most
reliable  approach, PCA on log2-1 on the  abundance  data, , which retains the
increased  impact of high  abundance  plant species on the analysis,  although
damped due to the log transformation, shows a clear species separation. Symbols and color codes as described above.

```{r fig.width=10,fig.height=10,fig.cap="PCA on log2 plant family abundance data"}
par(omi=c(1,0.6,0.6,0.3))
sbi_pca_pairs(pca,n=3,pch=pchs,col=cols,groups=spec,legend=FALSE,cex=3)
trachy$bottom_legend(labels=names(colors)[c(2,4,3,1)],col=unlist(colors)[c(2,4,3,1)],cex=1.2)
```

## Loadings of the first component

As it is visible that the first component separates the only species, living in rainforest. 
We will analyze the loadings of that component.

```{r results="asis"}
knitr::kable(head(pca$rotation[order(pca$rotation[,1],decreasing=TRUE),1:4]))
```

We can as well create a biplot to visualize those main loadings. for the first two compments:

```{r fig=TRUE,results="asis",fig.width=6,fig.height=6}
plot(pca$x[,1:2],col=cols[spec],pch=pchs,cex=2)
grid()
loadings=pca$rotation
fac=20
loadings=loadings[order(abs(loadings[,1]),decreasing=TRUE),]
arrows(0,0,loadings[1:5,'PC1']*fac,loadings[1:5,'PC2']*fac,
        length=0.1,angle=20,col='black')
text(loadings[1:5,'PC1']*fac*1.2,loadings[1:5,'PC2']*fac*1.2,
       substr(rownames(loadings)[1:5],1,5),col='black',font=2)
loadings=loadings[order(abs(loadings[,2]),decreasing=TRUE),]       
arrows(0,0,loadings[1:5,'PC1']*fac,loadings[1:5,'PC2']*fac,
        length=0.1,angle=20,col='blue')
text(loadings[1:5,'PC1']*fac*1.2,loadings[1:5,'PC2']*fac*1.2,
       substr(rownames(loadings)[1:5],1,5),col='blue',font=2)
```

**Figure: Biplot for PCA of plant family abundance  data.** The arrows show the loadings
of the original variables. For instance the Fagaceae, the Araliaceae and the Lentibulariaceae families are highly abundant in the T. crepusculus species.

Let's now test that these are significant higher in T. crepusculus, first we compare the mean:



```{r fig=TRUE,fig.width=10,fig.height=10}
library(lattice)
dlog2=as.matrix(log2(mdata[,6:(ncol(mdata))]+1))
dim(dlog2)
length(spec)
bwdata=data.frame(y=c(dlog2[,'Fagaceae'],dlog2[,'Araliaceae'],dlog2[,'Lentibulariaceae'],dlog2[,'Pentaphylacaceae']),
    family=as.factor(rep(c('Fagaceae','Araliaceae','Lentibulariaceae','Pentaphylacaceae'),each=nrow(dlog2))),
    species=as.factor(spec))
summary(bwdata)    
bwplot(y ~ species | family,data=bwdata)    
saveRDS(cbind(spec=spec,mdata),file="~/mdata.RDS")
```
    
And now the testing:

```{r}
for (fam in c('Fagaceae','Araliaceae','Pentaphylacaceae','Lentibulariaceae')) {
    print(fam)
    rkt=kruskal.test(as.vector(dlog2[,fam]), as.factor(spec))
    print(rkt)
    print(sbi_epsilon_squared(as.vector(dlog2[,fam]), as.factor(spec)))
    print(pairwise.wilcox.test(as.vector(dlog2[,fam]), as.factor(spec)))
}   
```

## Loadings of the second component

As it is visible that the second component separates the southern species, living as rainforest species on limbstone. 
We will analyze the loadings of that component.

```{r results="asis"}
knitr::kable(head(pca$rotation[order(pca$rotation[,2],decreasing=TRUE),1:4]))
```

```{r fig=TRUE,fig.width=10,fig.height=10}
library(lattice)
dlog2=as.matrix(log2(mdata[,6:(ncol(mdata))]+1))
dim(dlog2)
length(spec)
bwdata=data.frame(y=c(dlog2[,'Combretaceae'],dlog2[,'Phyllanthaceae'],
    dlog2[,'Vitaceae'],dlog2[,'Sapindaceae']),
    family=as.factor(rep(c('Combretaceae','Phyllanthaceae','Vitaceae','Sapindaceae'),each=nrow(dlog2))),
    species=as.factor(spec))
summary(bwdata)    
bwplot(y ~ species | family,data=bwdata)    
```
    
And now the testing:

```{r}
for (fam in c('Combretaceae','Phyllanthaceae','Vitaceae','Sapindaceae')) {
    print(fam)
    rkt=kruskal.test(as.vector(dlog2[,fam]), as.factor(spec))
    print(rkt)
    print(sbi_epsilon_squared(as.vector(dlog2[,fam]), as.factor(spec)))
    print(pairwise.wilcox.test(as.vector(dlog2[,fam]), as.factor(spec)))
}   
```

## Family based comparisons

We will  loop  over  all  plantfamilies  and  compute  the  kruskal  test  and
thereafter  a  Wilcox  test of Tcre (C)  against  all the  others  (DGH),  the
Tcre+Tger (CG) against Tdel+That (DH).

Let's first create these classes:

```{r}
suppressWarnings({
R=trachy_summary_data(mdata,spec)
})
```

Let's now plot the  significant  comparisons  between  the habitat  groups and
visualize the top ten highest effect sizes of Wilcoxon R. We display the value as negative if the second group has larger mean values then the first. Another requirement is that at least for one species the maximum percentage value of that value must be in one setting, dry or wet season above 1%.

```{r label=effsize,fig=TRUE,fig.width=6,fig.height=12,out.width=800}
par(mfrow=c(2,1),mai=c(0.3,0.9,0.9,0.2))
sset=R[,c('max','wt.habitat.pvalue','wt.habitat.r','C','DGH')]
idx=which(sset[,'C']<sset[,'DGH'])
sset$wt.habitat.r[idx]=-sset$wt.habitat.r[idx]
idx=which(sset$wt.habitat.pvalue<0.05 & sset$max>1)
sset=sset[idx,]
rownames(sset)
rn1=rownames(sset)
print(length(rownames(sset)))
print(sset[1:10,])
sset=sset[order(abs(sset$wt.habitat.r),decreasing=TRUE),]
trachy_es_plot(sset$wt.habitat.r[1:10],substr(rownames(sset)[1:10],1,4),
    main="Tcre vs Tdel+Tger+That",
    ylab="Wilcoxon Effect Size (r)")
sset=R[,c('max','wt.habspec.pvalue','wt.habspec.r','CG','DH')]
idx=which(sset[,'CG']<sset[,'DH'])
sset$wt.habspec.r[idx]=-sset$wt.habspec.r[idx]
idx=which(sset$wt.habspec.pvalue<0.05 & sset$max>1)
sset=sset[idx,]
sset=sset[order(abs(sset$wt.habspec.r),decreasing=TRUE),]
rownames(sset)
print(length(rownames(sset)))
print(sset[1:10,])
trachy_es_plot(sset$wt.habspec.r[1:10],
    substr(rownames(sset)[1:10],1,4),
    main="Tcre+Tger vs Tdel+That",
    ylab="Wilcoxon R Effect Size")
rn2=rownames(sset)    
nset=as.matrix(R[,c('Tcre','Tdel','Tger','That')])
nset=round(prop.table(nset,2)*100,3)
rownames(nset)=rownames(R)
nset=nset[union(rn1,rn2),]
saveRDS(list(Results=R,Percentages=nset,MainData=mdata),file="~/.R.RDS")

```

Let's describe the columnnames of the final Excel file:

- 'max' - the maximumn value for all species in percent
- 'kt.pvalue' - the p-value for the Kruskal-Wallis test
- 'epsq' - the epsilon squared effect size for the Kruskal-Wallis test
- 'wt.habitat.pvalue' - the p-value of a Wilcoxon test for the comparison of Tcre vs Tdel+Tger+That
- 'wt.habitat.r' - the Wilcoxon effect size (r) for that Wilcoxon test for the comparison of Tcre vs Tdel+Tger+That
- 'wt.habspec.pvalue' - the p-value of a Wilcoxon test for the comparison of Tcre+Tger vs Tdel+That
- 'wt.habspec.r' - the Wilcoxon effect size (r) for that Wilcoxon test for the comparison of Tcre+Tger vs Tdel+That
- 'Tcre','Tdel','Tger','That' - the mean values for these species
- 'C' - the mean value for 'Tcre'
- 'DGH' - the mean values for the three species  'Tdel','Tger' and 'That' taken together
- 'CH' - the mean value for 'Tcre' and 'Tger' taken together
- 'DH' - the mean values for the three species  'Tdel', and 'That' taken together
- 'PC1' - the loadings/rotations for the first component of the original data log2(x+1) transformed
- 'PC2' - the loadings/rotations for the second component of the original data log2(x+1) transformed

## Venn

```{r fig=TRUE,width=8,height=8,out.width=600}
par(mai=c(0.1,0.1,0.1,0.1))
sbi_venn(rn1,rn2,vars=c("C vs DGH","CG vs DH"),cex=1.2)
intersect(rn1,rn2)
```

## Genus Data Analysis

```{r}
genus=trachy$tdata("genus")
dim(genus)
genus[1:6,1:12]
```

### MDS of Plant Genus

```{r fig=TRUE,fig.width=8,fig.height=8,out.width=800}
par(mai=c(0.8,0.8,0.5,0.2))
M=t(as.matrix(genus[,3:ncol(mdata)]))
dim(M)
M[1:6,1:5]
col=rep("black",nrow(M))
col[1:5]="red"
rownames(M)=substr(rownames(M),1,4)
trachy$mds_plot(M,method="spearman",col.labels=col,grid=FALSE)
abbrev.gn=data.frame(abbrev=substr(colnames(genus)[3:(ncol(genus))],1,4),
    genus=colnames(genus)[3:(ncol(genus))])
```

**Figure  MDS of plant  families:**  Multidimensional  Scaling of the Spearman
correlation  based  distances ( D=(1-rho)/2 ) for the plant families. The five
most abundant  species,  Moraceae  (Mor),  Fagaceae (Fag),  Sapotaceae  (Sap),
Fabaceae   (Fab)  and  Rubiaceae   (Rub)  are  shown  in  red.  Full  list  of
abbreviations is given in the supplementary materials.

### Attach Meta Data

```{r}
dim(genus)
neta=meta=unique(meta[,c("habitat","species","seasonal","myear")])
gdata=merge(neta,genus,by.x=c("myear","species"),by.y=c("month","species"))
dim(gdata)
gdata[1:5,1:12]
```

### Season Medians

Let's calculate the season medians. For this we first merge the data:

```{r}
agg=with(gdata,aggregate(gdata[,5:(ncol(gdata))],
    by=list(seasonal,species),median))
colnames(agg)[1]="Season"
colnames(agg)[2]="Species"
agg[,1:10]
```

Let's tabulate this:

```{r results="asis"}
library(knitr)
aggr=agg[,1:12]
aggr[,3:12]=round(aggr[,3:12],1)
aggr[,2]=gsub(" ","",aggr[,2])
colnames(aggr)=substr(colnames(aggr),1,6)
colnames(aggr)[1]="Season"
colnames(aggr)[2]="Species"
knitr::kable(aggr)
```

### Visualization of Plant genus similarities using MDS

```{r fig=TRUE,fig.width=8,fig.height=8,out.width=800}
par(mai=c(0.8,0.8,0.5,0.2))
M=t(as.matrix(gdata[,5:ncol(gdata)]))
dim(M)
M[1:6,1:5]
col=rep("black",nrow(M))
col[1:5]="red"
rownames(M)=substr(rownames(M),1,4)
trachy$mds_plot(M,method="spearman",col.labels=col,grid=FALSE)
```

**Figure  MDS of plant genus:**  Multidimensional  Scaling of the Spearman
correlation  based  distances ( D=(1-rho)/2 ) for the plant families. The five
most abundant  genus,  Ficus  (Ficu),  Sideroxylon (Side),  Phanera  (Phan),
Celtis   (Celt)  and  Dalbergia (Dalb)  are  shown  in  red.  Full  list  of
abbreviations is given in the supplementary materials.

### MDS for genus data of samples

We will now use the aggregated data.

```{r fig.cap="MDS genus data with Spearman correlation as distance measure",fig.width=8,fig.height=8,out.width=800}
specg = gsub("T.([a-z]{3}).+","T\\1",gdata$species)
colsg=unlist(colors[specg])
head(colsg)
pchsg=c(18,20)[as.numeric(as.factor(gdata$seasonal))]
S=as.matrix(gdata[,5:ncol(gdata)])
dim(S)
trachy$mds_plot(S,
    method="spearman",points=TRUE,pch=pchsg,cex=3,col=colsg,row.labels=FALSE,grid=FALSE)
legend('bottomleft',legend=names(colors)[c(2,4,1,3)],fill=unlist(colors)[c(2,4,1,3)],cex=1.2)
legend('bottomright',legend=c("Dry","Wet"),pch=c(18,20),cex=1.2,pt.cex=2)
```

**Figure  MDS of Trachypithecus samples:**  Multidimensional  Scaling of the Spearman
correlation  based  distances ( D=(1-rho)/2 ) for the average values for each month based of the plant family data.  Every species has up to 12 datapoints, 

### Averages of the MDS

```{r fig=TRUE,fig.width=8,fig.height=8,out.width=800}
trachy$mds_plot(S,
    method="spearman",points=TRUE,
    pch=15,cex=3.2,col=colsg,row.labels=FALSE,grid=FALSE,group.mean=TRUE)
legend('bottomleft',legend=names(colors)[c(2,4,1,3)],fill=unlist(colors)[c(2,4,1,3)],cex=1.2)    
```


### MDS plot for every species

```{r fig=TRUE,fig.width=10,fig.height=10,out.width=1000}
par(mfrow=c(2,2),mai=c(0.7,0.7,0.6,0.2),omi=c(0.5,0.2,0.1,0.1))
options(warning=-1)
sp = unique(agg$Species)
sp = sp[c(2,4,3,1)]
x = 0
for (species in sp) {
    x=x+1
    mdat=gdata[gdata$species==species,5:ncol(gdata)]
    idx=which(apply(mdat,2,sum)>5)
    mdat=mdat[,idx]
    a=agg[agg$Species==species,colnames(mdat)]
    colnames(mdat)=substr(colnames(mdat),1,4)
    lf=log2(a[1,]/a[2,])
    col=rep("black",length(lf))
    col[lf < -1] = "blue"
    col[lf >  1] = "orange"
    xlab="Dim 1"
    ylab="Dim 2"
    if (x %in% c(1,2)) {
        xlab=""
    }
    if (x %in% c(2,4)) {
        ylab=""
    }
    if (x == 1) {
        par(mai=c(0.5,0.7,0.6,0.1))
    }
    if (x == 2) {
        par(mai=c(0.5,0.6,0.6,0.2))
    }
    if (x == 3) {
        par(mai=c(0.7,0.7,0.4,0.1))
    }
    if (x == 4) {
        par(mai=c(0.7,0.6,0.4,0.2))
    }
    trachy$mds_plot(t(mdat),method="spearman",grid=FALSE,main=species,col.labels=col,
        xlab=xlab,ylab=ylab)
}
trachy$bottom_legend(
    labels=c("dry","same","wet"),col=c("orange","black","blue"),cex=1.5)
```

**Figure  MDS of plant genus foreach Trachypithecus species:**  Multidimensional  Scaling of the Spearman
correlation  based  distances ( D=(1-rho)/2 ) forthe plant families.  Families
where the  log2-changes  between  wet and dry season are lower than -1 (plants
found more often in wet  season)  are shown in blue color, and plant  families
found  more  often in dry  season  (log2-foldchange  > 1) are shown in orange.
Families not having a log2-foldchange between -1 and 1 are shown in black.

### PCA

```{r fig.width=10,fig.height=10,fig.cap="PCA on plant genus abundance data"}
par(mfrow=c(2,2))
pcg=prcomp(gdata[,5:(ncol(gdata))])
pca.plot(pcg,col=colsg,pch=pchsg,cex=3,main="raw data")
pcg=prcomp(scale(gdata[,5:(ncol(gdata))]))
pca.plot(pcg,col=colsg,pch=pchs,cex=3,main="scaled data")
pcg=prcomp(scale(log2(gdata[,5:(ncol(gdata))]+1)))
pca.plot(pcg,col=colsg,pch=pchsg,cex=3,main="scaled log-data")
pcg=prcomp(log2(gdata[,5:(ncol(gdata))]+1))
pca.plot(pcg,col=colsg,pch=pchsg,cex=3,main="log-data")
```

**Figure: PCA of plant genus abundance  data.** Shown are the results for the
raw, data, z-score scaled data, log2+1 transformed z-score data and log2-data.
Regardless of the data normalization all plots show a clear species separation
and a less  obvious  separation  of the  seasons  for each  species.  The most
reliable  approach, PCA on log2-1 on the  abundance  data, , which retains the
increased  impact of high  abundance  plant species on the analysis,  although
damped due to the log transformation, shows a clear species separation. Symbols and color codes as described above.


```{r fig.width=10,fig.height=10,fig.cap="PCA on log2 plant genus abundance data"}
par(omi=c(1,0.6,0.6,0.3))
sbi_pca_pairs(pcg,n=3,pch=pchsg,col=colsg,groups=spec,legend=FALSE,cex=3)
trachy$bottom_legend(labels=names(colors)[c(2,4,3,1)],col=unlist(colors)[c(2,4,3,1)],cex=1.2)
```


## Supplementary Materials


### PCA of log2 normalized family data:

```{r fig.width=10,fig.height=10,fig.cap="PCA on log2 plant family abundance data"}
par(omi=c(1,0.6,0.6,0.3))
sbi_pca_pairs(pca,n=5,pch=pchs,col=cols,groups=spec,legend=FALSE,cex=3)
trachy$bottom_legend(labels=names(colors)[c(2,3,4,1)],col=unlist(colors)[c(2,3,4,1)],cex=1.2)
```

### PCA of log2 normalized genus data:

```{r fig.width=10,fig.height=10,fig.cap="PCA on log2 plant genus abundance data"}
par(omi=c(1,0.6,0.6,0.3))
sbi_pca_pairs(pcg,n=5,pch=pchsg,col=colsg,groups=specg,legend=FALSE,cex=3)
trachy$bottom_legend(labels=names(colors)[c(2,3,4,1)],col=unlist(colors)[c(2,3,4,1)],cex=1.2)
```


```{r results="asis"}
abbrev.df=abbrev.df[order(abbrev.df$abbrev),]
knitr::kable(abbrev.df)
```

```{r results="asis"}
abbrev.gn=abbrev.gn[order(abbrev.gn$abbrev),]
knitr::kable(abbrev.gn)
```




